---
title: "glycoTraitR Quick Start"
author: "Bingyuan Zhang"
date: "`r Sys.Date()`"
abstract: >
  glycoTraitR provides a streamlined workflow for analyzing glycan
  heterogeneity in glycoproteomics GPSM data generated by pGlyco3 or
  Glyco-Decipher. The package summarizes glycan structures into
  biologically interpretable traits, enabling downstream statistical
  analysis and visualization.
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::knitr}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Introduction

glycoTraitR provides a streamlined workflow for extracting, summarizing, and analyzing glycan structural traits from glycoproteomics GPSM data generated by **pGlyco3** or **Glyco-Decipher**. Rather than focusing on individual glycoforms, glycoTraitR summarizes glycan structures into biologically interpretable composition- and structure-based traits at the site or protein level.

In this quick start guide, we demonstrate how to:

- import GPSM results generated by pGlyco3 or Glyco-Decipher
- parse glycan structures represented as pGlyco3 strings or WURCS 2.0
- perform differential trait analysis using a permutation-based two-sample test
- visualize trait-level changes in glycan heterogeneity

## Installation

The glycoTraitR package can be installed using **BiocManager**. 

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("glycoTraitR")
```

After installation, we load the package into the current R session. All functions used in the following examples are provided by *glycoTraitR*.

```{r load}
library(glycoTraitR)
```

# Quick Start

This section demonstrates a typical *glycoTraitR* workflow using GPSM data
generated by **pGlyco3**. Equivalent import code for **Glyco-Decipher**
results is also provided.

## 1. Load example data

For reproducibility, this vignette uses a toy dataset bundled with
*glycoTraitR*. The commented code below illustrates how to download and
import the corresponding full public datasets from **Zenodo**.

```{r load-example-data}
# Option 1: Download and import the pGlyco3 result
# pglyco_url <- paste0(
#   "https://zenodo.org/records/21386154/files/",
#   "pGlycoDB-GP-FDR-Pro-Quant-Site.txt?download=1"
# )
# download.file(
#   pglyco_url,
#   destfile = "pGlyco3_GPSM.txt",
#   mode = "wb"
# )
# gpsm <- read_pGlyco3_gpsm("pGlyco3_GPSM.txt")

# Option 2: Download and import the Glyco-Decipher result
# decipher_url <- paste0(
#   "https://zenodo.org/records/21386154/files/",
#   "decipher_result.zip?download=1"
# )
# download.file(
#   decipher_url,
#   destfile = "decipher_result.zip",
#   mode = "wb"
# )
# unzip("decipher_result.zip")
# gpsm <- read_decipher_gpsm("decipher_result")

# Download the sample metadata used by both examples
# meta_url <- paste0(
#   "https://zenodo.org/records/21386154/files/",
#   "meta_mcp_2022_100433.rds?download=1"
# )
# download.file(
#   meta_url,
#   destfile = "meta_example.rds",
#   mode = "wb"
# )
# meta <- readRDS("meta_example.rds")

# Load the bundled pGlyco3 toy dataset and sample metadata
gpsm <- readRDS(system.file("extdata", "gpsm_toyexample.rds", package = "glycoTraitR"))
meta <- readRDS(system.file("extdata", "meta_toyexample.rds", package = "glycoTraitR"))

head(gpsm)
head(meta)
```

The `gpsm` object contains glycopeptide-spectrum matches, including glycan structures encoded in pGlyco3 format.  
The `meta` object provides sample-level annotations that will later be used to define experimental groups.

## 2. Compute heterogeneity scores and test group differences

After importing the GPSM data, we use `analyze_hscore_changes()` to perform
the main analysis. This wrapper function parses the identified glycan
structures, extracts glycan traits, computes sample-level heterogeneity
scores, and tests for differences between experimental groups.

By default, heterogeneity is evaluated at both the glycosylation-site and
protein levels. In this example, samples labeled as *Normal* and *Symptomatic* are compared according to the `Diagnosis` column in the metadata. Because the bundled toy dataset was generated by pGlyco3, we set `from = "pGlyco3"`. For data imported with `read_decipher_gpsm()`, use `from = "decipher"` instead.

```{r analyze-hscore}
set.seed(123)
res <- analyze_hscore_changes(
  gpsm = gpsm,
  from = "pGlyco3", 
  motifs = NULL,
  meta = meta,
  group_col = "Diagnosis",
  group_levels = c("Normal", "Symptomatic"),
  B = 1000,
  min_samples = 3
)
```

The `gpsm` argument supplies the imported GPSM table, while `from` specifies the search-engine format. The `meta` table links sample identifiers to their corresponding experimental groups. The `B` argument controls the number of label permutations used to construct the null distribution, and `min_samples` specifies the minimum number of samples required for a trait–feature combination to be tested.

The returned object contains the results from both site- and protein-level analyses. Each row represents a glycan trait evaluated for a specific glycosylation site or protein and includes the estimated group difference and its associated statistical significance.

## 3. Inspect significant heterogeneity changes

The complete result table can be filtered to identify trait–feature combinations showing evidence of altered heterogeneity between the two groups. For example, the following code displays results with an unadjusted p-value below 0.05.

```{r inspect-hscore}
res[res$pval < 0.05, ]
```

These results represent glycan traits whose heterogeneity differs between the *Normal* and *Symptomatic* groups at either the glycosylation-site or protein level. The corresponding analysis level can be identified from the relevant result column.

## 4. Visualize heterogeneity changes

The analysis results can be visualized using a heatmap and a volcano plot.

```{r visualize-hscore}
plot_hscore_heatmap(res)
plot_hscore_volcano(res)
```

The heatmap summarizes group-associated heterogeneity patterns across
selected trait–feature combinations, facilitating comparisons among
proteins, glycosylation sites, and glycan traits.

The volcano plot displays the magnitude of each heterogeneity change
together with its statistical significance. Trait–feature combinations
with larger absolute group differences appear farther from zero on the
horizontal axis, whereas those with stronger statistical evidence appear
higher on the vertical axis.

# Define Your Own Glycan Motifs

In addition to the pre-defined glycan traits, glycoTraitR allows users to define custom glycan motifs of biological interest. These motifs are detected within glycan structures and can be incorporated directly into downstream trait analyses.

In this vignette, we will:

-  Parse glycan structures into graph trees 
-  Visualize full glycans & motifs 
-  Perform **subgraph isomorphism** matching using *igraph* functions 
-  Count occurrences of motif patterns

```{r load-igraph, message=FALSE}
library(igraph)
```

-----

## 1. Parse and plot glycan trees

We first parse glycan structures into tree representations.  
Internally, glycans are represented as graphs, where nodes correspond to monosaccharides and edges represent glycosidic linkages.

```{r example-pGlyco3}
# Example: parsed glycan trees from existing GPSM data
glycans <- lapply(gpsm$GlycanStructure, glycoTraitR:::pGlyco3_to_tree)
```

We use the function `build_glycan_igraph()` to visualize an example of glycan structure.

```{r visualize-tree}
# visualize example structures
igraph_obj <- build_glycan_igraph(glycans[[1]])
plot_glycan_tree(igraph_obj)
```

This visualization highlights the branching structure and residue composition of a glycan, which is essential for understanding motif topology.


------------------------------------------------------------------------

## 2. A Simple User Motif (Example 1)

This example demonstrates how to define and detect a simple glycan motif consisting of three hexose (H) residues connected in series.

Here we define a **linear trisaccharide motif**:

-   H — H — H
-   Edges: `"a-b"`, `"b-c"`

```{r example1}
#### Example 1 ####
# (1) Select a glycan from pGlyco3 data
tree_full <- glycans[[2]]
g_full <- build_glycan_igraph(tree_full)

# (2) Define motif manually
motif <- list(
  node = c("H", "H", "H"),
  edge = c("a-b", "b-c")
)
g_motif <- build_glycan_igraph(motif)

# (3) Perform subgraph matching
type_levels <- union(
    unique(V(g_full)$type),
    unique(V(g_motif)$type)
)
n_match <- count_subgraph_isomorphisms(
  g_motif, g_full,
  method = "vf2",
  vertex.color1 = factor(V(g_full)$type, levels=type_levels),
  vertex.color2 = factor(V(g_motif)$type, levels=type_levels)
)

# (4) Visualize
par(mfrow = c(1, 2))
plot_glycan_tree(g_motif)
plot_glycan_tree(g_full)
print(paste(n_match, "motif(s) found in the glycan"))
```

The reported match count indicates how many times the specified motif occurs within the full glycan structure.  

------------------------------------------------------------------------

## 3. Motif Containing Fucose (Example 2)

More complex motifs can also be defined, including branched structures and specific monosaccharide types such as fucose (F).  

This example defines a **branched motif** containing Fucose:

```{r example2}
#### Example 2 ####
# (1) Select a glycan
tree_full <- glycans[[6]]
g_full <- glycoTraitR::build_glycan_igraph(tree_full)

# (2) Symbolic motif definition
motif <- list(
  node = c("H", "N", "F"),
  edge = c("a-b", "b-c")
)
g_motif <- glycoTraitR::build_glycan_igraph(motif)

# (3) Subgraph matching
type_levels <- union(
  unique(V(g_full)$type),
  unique(V(g_motif)$type)
)

n_match <- igraph::count_subgraph_isomorphisms(
  g_motif,
  g_full,
  method = "vf2",
  vertex.color1 = factor(V(g_full)$type, levels = type_levels),
  vertex.color2 = factor(V(g_motif)$type, levels = type_levels)
)

# (4) Visualize motif vs. full glycan
par(mfrow = c(1, 2))
plot_glycan_tree(g_motif)
plot_glycan_tree(g_full)
print(paste(n_match, "motif(s) found in the glycan"))
```

------------------------------------------------------------------------

## 4. Integrating Motifs with glycoTraitR (for Trait Computation)

Multiple user-defined motifs can be supplied simultaneously.  
Once defined, these motifs are treated as additional glycan traits and are incorporated into the trait computation workflow.

```{r udmotifs}
user_motifs <- list(
  LinearH3 = list(
    node = c("H", "H", "H"),
    edge = c("a-b", "b-c")
  ),
  FucBranch = list(
    node = c("H", "N", "F"),
    edge = c("a-b", "b-c")
  )
)
```

Each user-defined motif can be input as an option.

```{r se-show}
set.seed(123)
res <- analyze_hscore_changes(
  gpsm = gpsm,
  from = "pGlyco3",
  motifs = user_motifs,
  meta = meta,
  group_col = "Diagnosis",
  group_levels = c("Normal", "Symptomatic"),
  B = 1000,
  min_samples = 3
)
res[res$trait == "LinearH3" & res$pval < 0.05, ]
```

Your motif counts will automatically appear as **additional glycan traits** in the trait matrices.

------------------------------------------------------------------------

## Conclusion

In this vignette, we demonstrated how to define custom glycan motifs, detect their occurrences within glycan structures, and incorporate them into the glycoTraitR trait analysis workflow.

```{r session-info, echo=FALSE}
sessionInfo()
```

